# Reproducible Research: Peer Assessment 1

Insert introductory text here

## Load the R packages to be used during the session

```{r}
library(ggplot2)
library("plyr") 
```

## Load and preprocess the data

```{r}
unzip("activity.zip")  
data <- read.csv("activity.csv", stringsAsFactors=FALSE)
data$date <- strptime(data$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Develop histogram graph of steps taken per day

```{r}
sums <- ddply(data, 
      .(date), 
      summarise, 
      sum_steps = sum(steps, na.rm = TRUE),
      sum_int = sum(interval, na.rm=TRUE))

m <- ggplot(sums, aes(x=sum_steps)) + 
      geom_histogram(binwidth = 1000,
                     aes(fill = ..count..)) +
      labs(x = "Total Steps Taken Per Day", 
            y = "Count of Steps", 
            title = "Total Steps Taken Per Day")

print(m)

```

Calculate mean and median steps takend per day

```{r}
mean(sums$sum_steps)
median(sums$sum_steps)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
